Simplified and Verified: A Second Look at a Proof-Producing Union-Find Algorithm

### 2024-09-17 paper/CPP/paper.tex
Continue

\usepackage{tikz}
\usepackage{pgfplots}

@@

\usepackage[acronym]{glossaries}
\newacronym{adt}{ADT}{Abstract Data Type}
\newacronym{uf}{UF}{Union-Find}
\newacronym{ufe}{UFE}{Union-Find-Explain}

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes,defines},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

% Notation
\newcommand{\TransP}{\bigtriangledown}

@@

\title{Simplified and Verified: A Fresh Look at a Proof-Producing Union-Find Algorithm}
\author{Lukas Stevens}
\orcid{0000-0003-0222-6858}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{lukas.stevens@in.tum.de}

@@

\section{Introduction}
\begin{itemize}
  \item Based on \cite{congcl_proofs}
\end{itemize}
\subsection{Contributions}

\subsection{Related Work}
\begin{itemize}
  \item Congruence closure in Coq by Corbineau
\end{itemize}

@@

We use \lstinline!'a list! to describe the type of lists, which are constructed using the empty list \lstinline![]! constructor or the infix cons constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
\todo[inline]{Indexing operator}
\todo[inline]{Relation type synonym, Field of a relation}
\todo[inline]{Record notation}

@@

According to \citeauthor{unionfind_tarjan}~\cite{unionfind_tarjan}, this data structure can be implemented as a forest of rooted trees where each tree represents an equivalence class.
The edges of the trees are directed towards their root, which is the representative of the corresponding equivalence class.
To keep this invariant, we initialise the forest with $n$ nodes but without any edges and for every \lstinline!union! of $a_i$ and $a_j$ we add a directed edge from \lstinline!rep_of $l$ $a_i$! to \lstinline!rep_of $l$ $a_j$! to the forest.

@@

\todo[inline]{Mention optimisations?}

\subsection{In Isabelle/HOL}
The \acrshort{uf} algorithm is formalised in Isabelle/HOL by \citeauthor{unionfind_isabelle}~\cite{unionfind_isabelle}.
The code can be found in an entry~\cite{union_find_afp} of the \emph{Archive of Formal Proofs} (AFP).\footnote{The code is in the theory file \texttt{Examples/Union\_Find.thy}.}
Later, \citeauthor{union_find_imp_hol}~\cite{union_find_imp_hol} also formalised the running time.\todo{Check citations}

The formalisation defines a function \lstinline|rep_of|, which, as described in the previous section, follows the parent pointers until we arrive the root, where the parent pointer is self-referential.
\begin{lstlisting}[columns=fullflexible]
function rep_of :: nat list $\Rightarrow$ nat $\Rightarrow$ nat where
rep_of l a =
  if l ! a = a then a else rep_of l (l ! a)
\end{lstlisting}
Looking closely at this definition, we see that this function is only well-defined for some inputs \lstinline|l| and \lstinline|a|:
for every element \lstinline|a < length l|, its parent must be in the list, i.e.\ we must have \lstinline|l ! a < length l|, and the parent pointers must be cycle-free in order for the function to terminate.
Functions in Isabelle/HOL must be total, so the \lstinline|function| command introduces a constant
\begin{lstlisting}
rep_of_dom :: (nat list $\times$ nat) $\Rightarrow$ bool  
\end{lstlisting}
that characterises the inputs for which \lstinline|rep_of| terminates.
Then, it adds \lstinline|rep_of_dom (l, a)| as a premise to the defining equation of \lstinline|rep_of|. 
The intuition above is cast into a predicate \lstinline|ufa_invar| that defines such well-formed lists \lstinline|l|.
\begin{lstlisting}
ufa_invar l = $\forall$a < length l. rep_of_dom (l, a)
              $\land$ l ! a < length l
\end{lstlisting}

Building on this formalisation, we encapsulate this \acrshort{uf} data structure into an \acrfull{adt} \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the \acrshort{adt} using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to, and
  \item \lstinline!ufa_union uf x y! returns a \acrshort{uf} data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! with \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:\todo{Spacing between lemmas,defs}
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

lemma ufa_$\alpha$ (ufa_union uf x y) =
  per_union (ufa_$\alpha$ uf) x y
\end{lstlisting}
where \lstinline!per_union R x y!\todo{Definition with \lstinline|trancl|?} is defined as the equivalence relation that results from merging the respective equivalence classes in the relation \lstinline|R| that \lstinline!x! and \lstinline!y! belong to.

But what happens if \lstinline!x! or \lstinline!y! is not an element of the partial equivalence relation \lstinline|R|?\todo{Define element of per.}
In that case, the equivalence relation is unchanged, which means that \lstinline|per_union R x y = R|.
This, however, can be seen as a misuse of the \acrshort{uf} data structure, since we initialise it with a fixed set of elements \lstinline|A| and expect the user to only work with these elements.
Therefore, we introduce the following definitions that characterise valid union(s) with regard to the initial set \lstinline|A|.
\begin{lstlisting}
definition valid_union uf a b =
  a $\in$ Field (ufa_$\alpha$ uf) $\land$ b $\in$ Field (ufa_$\alpha$ uf)
definition valid_unions uf us =
  $\forall$(x, y) $\in$ set us. valid_union uf x y
\end{lstlisting}

\section{Simple Certifying Union-Find Algorithm}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2 : (x, z)|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
Building on the \acrshort{uf} \acrshort{adt} from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

@@

definition explain_partial us x y =
  if (x, y) $\in$ equivcl (set us) then Some (explain us x y) else None
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation. It operates on an arbitrary initial \acrshort{uf} data structure \lstinline|uf_init|.\label{fig:explain}}
\end{figure*}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with an instance \lstinline|uf| of the \acrshort{uf} \acrshort{adt} that results from applying \lstinline|us| as unions to an initial \acrshort{uf} data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
so we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.

At last, we are ready to implement the \lstinline|explain| operation as depicted in \autoref{fig:explain}.\todo{explain with Cons instead of snoc}
The algorithm works with the assumption that the given elements \lstinline|x| and \lstinline|y| are equal modulo \lstinline|us|.
Furthermore, we assume that each element of the initial data structure \lstinline|uf_init| is equal to itself
and that \lstinline|us| are valid unions with respect to \lstinline|uf_init|.
In formulae, we assume
\begin{itemize}
  \item \lstinline|ufa_$\alpha$ uf_init $\subseteq$ Id|, and
  \item \lstinline|valid_unions uf_init us|
\end{itemize}
in what follows.

@@

Otherwise, the list of unions must be equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming that the former holds --- the other case is similar --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the fact that \lstinline|a = b|, which we prove by assumption, we obtain \lstinline|x = y| by transitivity.
\end{itemize}

@@

\section{Efficient Certifying Union-Find Algorithm}
In the previous section, we developed a simple certifying \acrfull{ufe} data structure whose \lstinline|explain| operation iterates through a list of equalities,
identifying which of them are necessary in order to prove that its input values are equal.
Nevertheless, iterating through all equalities seems to be inefficient and indeed \citeauthor{congcl_proofs}~\cite{congcl_proofs} propose a more targeted approach.
Their approach adds two data structures to the list of equalities, which we collect into a record type \lstinline|ufe_ds|.
We call this collection of data structures the \acrshort{ufe} data structure.
For some fixed instance \lstinline|ufe_ds|, the purpose of the record members is as follows:
\todo[inline]{Introduce \lstinline|ufa_parent_of|, explain abstraction leakage. intuition here?. Name for the union-find-explain data structure?}
\begin{itemize}
  \item The list \lstinline|unions ufe_ds :: (nat $\times$ nat) list| contains the input equalities.
  \item The \acrshort{ufe} data structure \lstinline|uf_ds ufe_ds :: ufa| maintains the equivalence closure of the input equalities or, put differently, it is the result of applying \lstinline|unions ufe_ds| to an initial union find data structure.
    We require this to be a forest of rooted trees as introduced in \autoref{sec:uf_background}.
  \item The partial function \lstinline|au_ds ufe_ds :: nat $\Rightarrow$ nat option| associates an unique equality with each edge in the forest by indexing into \lstinline|unions ufe_ds|.
    In other words, there exists an \lstinline|i| such that \lstinline|au_ds ufe_ds x $\neq$ Some i| if \lstinline|x| has a parent in the forest.
    The associated union of the edge from \lstinline|x| to its parent is then \lstinline|unions ufe_ds ! i|.
\end{itemize}
For succinctness, we lift the operations \lstinline|ufa_rep_of|, \lstinline|ufa_parent_of|, and \lstinline|ufa_$\alpha$|\todo{Introduce \lstinline|ufa_parent_of|} from the record member \lstinline|uf_ds| to the record type \lstinline|ufe_ds|.
The lifted operations use the prefix \lstinline|ufe| instead of \lstinline|ufa|, so \lstinline|ufa_rep_of| is lifted to \lstinline|ufe_rep_of|, for example.

Similarly, we obtain the initial \acrshort{ufe} data structure by lifting \lstinline|ufa_init| while leaving the other two members empty as the following definition shows.
\todo[inline]{Introduce \lstinline|ufa_init|}
\begin{lstlisting}
definition ufe_init :: nat $\Rightarrow$ ufe_ds where
  ufe_init n = $\llparenthesis$ uf_ds = ufa_init n,
    au_ds = ($\lambda$_. None), unions = [] $\rrparenthesis$
\end{lstlisting}
Continuing with the union operation, we lift \lstinline|ufa_union| to the \acrshort{ufe} data structure.
For a union operation between \lstinline|x| and \lstinline|y|, we add \lstinline|(x, y)| to the list of unions and update the associated union of the representative of \lstinline|x| to \lstinline|(x, y)|.
We need to use the representative here because \lstinline|ufa_union| adds an edge from the representative of \lstinline|x| to that of \lstinline|y|.
As with the \acrshort{uf} data structure, we also define a function that applies a list of unions in consecutive order.
\begin{lstlisting}
fun ufe_union :: ufe_ds $\Rightarrow$ nat $\Rightarrow$ nat
              $\Rightarrow$ ufe_ds where
  ufe_union $\llparenthesis$ uf_ds = uf
            , au_ds = au, unions = u $\rrparenthesis$ x y =
    $\llparenthesis$ uf_ds = ufa_union uf x y
    , au_ds = au(ufa_rep_of uf x $\mapsto$ length u)
    , unions = u @ [(x, y)] $\rrparenthesis$

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes,positioning,patterns,patterns.meta}
\usepackage{pgfplots}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

@@

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,typedef,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive,record},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and,fixes,defines},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=\lst@ifdisplaystyle{fullflexible}\else{fixed}\fi,,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother
\newcommand{\lstinlinew}[2][]{\lstinline[#1]{#2}}
% This is a hack to allow escaping inside lstinline (https://tex.stackexchange.com/questions/43526/escaping-in-lstinline)
\usepackage{etoolbox}
\makeatletter
\patchcmd{\lsthk@TextStyle}{\let\lst@DefEsc\@empty}{}{}{\errmessage{failed to patch}}
\makeatother

@@

% Acronyms
\usepackage[acronym]{glossaries}
\newacronym{adt}{ADT}{abstract data type}
\newacronym{uf}{UF}{Union-Find}
\newacronym{ufe}{UFE}{Union-Find-Explain}
\newacronym{lca}{LCA}{lowest common ancestor}
\newacronym{afp}{AFP}{Archive of Formal Proofs}
\newacronym{smt}{SMT}{Satisfiability Modulu Theories}

% Notation
\newcommand{\TransP}{\bigtriangledown}
\newcommand{\NewestL}[3]{(#2 $\upharpoonleft$ #3)$_\text{#1}$}
\newcommand{\NewestR}[3]{(#2 $\upharpoonright$ #3)$_\text{#1}$}
\newcommand{\opunion}{union}
\newcommand{\opfind}{find}
\newcommand{\opexplain}{explain}

@@

\title{Simplified and Verified: A Second Look at a Proof-Producing Union-Find Algorithm}
\author{Lukas Stevens}
\orcid{0000-0003-0222-6858}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{lukas.stevens@in.tum.de}

\author{Rebecca Ghidini}
\orcid{0009-0009-8117-2659}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{rebecca.ghidini@tum.de}

@@

\section{Introduction}
The \acrfull{uf} data structure maintains the equivalence closure of a relation, where the relation is given as a sequence of pairs or, in terms of the \acrshort{uf} data structure, \opunion operations.
It is fundamental to efficiently implement well-known graph algorithms such as \citeauthor{mst}'s~\cite{mst} minimum spanning tree algorithm. 
There it tracks which vertices belong to the same connected component and are, in that sense, equivalent.
Beyond graph algorithms, its applicability extends to the domain of theorem proving as it routinely forms the basis of congruence closure algorithms, which are widely used by \acrfull{smt} solvers.
In order to increase their trustworthiness, current \acrshort{smt} solvers such as CVC5~\cite{cvc5}, E~\cite{eprover}, Vampire~\cite{vampire}, VeriT~\cite{verit}, and Z3~\cite{z3_proofs} are able to output detailed proofs if they determine their input formula to be unsatisfiable.

\begin{itemize}
  \item Based on \cite{congcl_proofs}
  \item Especially in the context of theorem proving certificates are useful
  \item Maybe cite order paper
\end{itemize}

\subsection{Contributions}
\begin{itemize}
  \item Formalisation of \acrshort{ufe} data structure
  \item Reduction of the complex recursion scheme to a simpler one that produces the same output
  \item Proof of soundness, completeness, and termination
  \item Abstract formalisation of union-find and union-find with size information hiding implementation details behind type definition
  \item Refinement of the \acrshort{ufe} to Imperative HOL to be able to export efficient code
\end{itemize}
\todo[inline]{We only show proof sketches, details in the formalisation}

\subsection{Related Work}
\begin{itemize}
  \item Union-Find as forest and union-by-size \cite{uf_by_size}
  \item Path compression \cite{uf_compress}
  \item Analysis of the run-time: $\mathcal{O}(n + m \alpha(m + n, n))$ for a sequence of at most $n - 1$ unions and $m$ finds \cite{uf_ub_improved}
  \item In Coq \cite{uf_coq} with time \cite{uf_coq_time}
  \item In Isabelle \cite{uf_isabelle} with time \cite{uf_isabelle} more recently viewed algebraically \cite{uf_isabelle_algebraic}.
\end{itemize}

@@

We use \lstinline!'a list! to describe the type of lists, which are constructed using the empty list \lstinline![]! constructor or the infix cons constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
\todo[inline]{Indexing operator}
\todo[inline]{Relation type synonym, Field of a relation}
\todo[inline]{Record notation}
\todo[inline]{Order on options}
\todo[inline]{Locale notation}
\todo[inline]{Lemma numbering}

@@

The data structure can be implemented as a forest of rooted trees where each tree represents an equivalence class.
The edges of the trees are directed towards their root, which is the representative of the corresponding equivalence class.
To keep this invariant, we initialise the forest with $n$ nodes but without any edges and for every \lstinline!union! of $a_i$ and $a_j$ we add a directed edge from \lstinline!rep_of $l$ $a_i$! to \lstinline!rep_of $l$ $a_j$! to the forest.

@@

\subsection{In Isabelle/HOL\label{sec:uf_hol}}
The \acrshort{uf} algorithm is formalised in Isabelle/HOL by \citeauthor{uf_isabelle}~\cite{uf_isabelle}.
The code can be found in an entry~\cite{uf_isabelle_afp} of the \acrfull{afp}.\footnote{The code is in the theory file \texttt{Examples/Union\_Find.thy}.}
Later, \citeauthor{uf_isabelle}~\cite{uf_isabelle} also formalised the running time.\todo{Check citations}

The formalisation defines a function \lstinline|rep_of|, which, as described in the previous section, follows the parent pointers until we arrive the root, where the parent pointer is self-referential.
\begin{lstlisting}[columns=fullflexible]
function rep_of :: nat list $\Rightarrow$ nat $\Rightarrow$ nat where
rep_of l a =
  if l ! a = a then a else rep_of l (l ! a)
\end{lstlisting}
Looking closely at this definition, we see that this function is only well-defined for some inputs \lstinline|l| and \lstinline|a|:
for every element \lstinline|a < length l|, its parent must be in the list, i.e.\ we must have \lstinline|l ! a < length l|, and the parent pointers must be cycle-free in order for the function to terminate.
Functions in Isabelle/HOL must be total, so the \lstinline|function| command introduces a constant
\begin{lstlisting}
rep_of_dom :: (nat list $\times$ nat) $\Rightarrow$ bool  
\end{lstlisting}
that characterises the inputs for which \lstinline|rep_of| terminates.
Then, it adds \lstinline|rep_of_dom (l, a)| as a premise to the defining equation of \lstinline|rep_of|. 
The intuition above is cast into a predicate \lstinline|ufa_invar| that defines such well-formed lists \lstinline|l|.
\begin{lstlisting}
definition ufa_invar l = $\forall$a < length l.
  rep_of_dom (l, a) $\land$ l ! a < length l
\end{lstlisting}
Building on this formalisation, we encapsulate this \acrshort{uf} data structure into an \acrfull{adt} \lstinline|ufa| by defining
\begin{lstlisting}
typedef ufa = {uf. ufa_invar uf}
\end{lstlisting}
By lifting operations on the underlying list to the \acrshort{adt} using Isabelle's lifting infrastructure~\cite{lifting_transfer}, we obtain the operations
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ :: ufa $\Rightarrow$ nat rel!,
  \item \lstinline!ufa_rep_of :: ufa $\Rightarrow$ nat $\Rightarrow$ nat!, and
  \item \lstinline!ufa_union :: ufa $\Rightarrow$ nat $\Rightarrow$ nat $\Rightarrow$ ufa!,
\end{itemize}
where \lstinline!nat rel = (nat $\times$ nat) set!.
The intended meaning of these operations is the following:
\begin{itemize}
  \item \lstinline!ufa_$\alpha$ uf! is the partial equivalence relation represented by \lstinline!uf!,
  \item \lstinline!ufa_rep_of uf x! is the representative of the equivalence class that \lstinline!x! belongs to, and
  \item \lstinline!ufa_union uf x y! returns a \acrshort{uf} data structure where the equivalence classes of \lstinline!x! and \lstinline!y! are merged.
    This is implemented by updating the underlying list at index \lstinline!rep_of l x! to \lstinline|rep_of l y|.
\end{itemize}
More rigorously, the above operations fulfil the following properties:
\begin{lstlisting}
lemma ufa_rep_of uf x = ufa_rep_of uf y $\longleftrightarrow$
  (x, y) $\in$ ufa_$\alpha$ uf

lemma ufa_$\alpha$ (ufa_union uf x y) =
  per_union (ufa_$\alpha$ uf) x y
\end{lstlisting}
where \lstinline!per_union R x y! is defined as the equivalence relation that results from merging the respective equivalence classes in the relation \lstinline|R| that \lstinline!x! and \lstinline!y! belong to.

But what happens if \lstinline!x! or \lstinline!y! is not an element of the partial equivalence relation \lstinline|R|?
In that case, the equivalence relation is unchanged, which means that \lstinline|per_union R x y = R|.
This, however, can be seen as a misuse of the \acrshort{uf} data structure, since we initialise it with a fixed set of elements \lstinline|A| and expect the user to only work with these elements.
Therefore, we introduce the following definitions that characterise valid union(s) with regard to this initial set.
\begin{lstlisting}
definition valid_union uf a b =
  a $\in$ Field (ufa_$\alpha$ uf) $\land$ b $\in$ Field (ufa_$\alpha$ uf)
definition valid_unions uf us =
  $\forall$(x, y) $\in$ set us. valid_union uf x y
\end{lstlisting}

\section{Simple Certifying Union-Find Algorithm}
\begin{figure}
  \begin{gather*}
    \inferrule*[right=Assm]{\hbox{\lstinline|i < length us|} \\ \hbox{\lstinline|us ! i = (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ AssmP i : (x, y)|}} \\[0.5em]
    \inferrule*[right=Refl]{ }{\hbox{\lstinline|us $\ \vdash_=\ $ ReflP x : (x, x)|}} \\[0.5em]
    \inferrule*[right=Sym]{\hbox{\lstinline|us $\ \vdash_=\ $ p : (x, y)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ SymP p : (y, x)|}} \\[0.5em]
    \inferrule*[right=Trans]{\hbox{\lstinline|us $\ \vdash_=\ $ p1 : (x, y)|} \\ \hbox{\lstinline|us $\ \vdash_=\ $ p2 (y, z)|}}{\hbox{\lstinline|us $\ \vdash_=\ $ p1 $\ \TransP\ $ p2 : (x, z)|}}
  \end{gather*}
  \caption{The data type \lstinline|eq_prf| of certificates and its corresponding system of inference rules $\vdash_=$.\label{fig:eq_prf}}
\end{figure}
\todo[inline]{Use equivalence instead of equality, Use equivalence operator instead of pairs, Maybe use cong symbol}
Building on the \acrshort{uf} \acrshort{adt} from the previous section, we now develop a simple \lstinline|explain| operation that,
for a given list of equations \lstinline|us|, takes two elements \lstinline|x| and \lstinline|y| and produces a certificate that \lstinline|x = y| modulo \lstinline|us|.
The certificate is given in terms of a data type \lstinline|eq_prf| with its corresponding system $\vdash_=$ of inference rules as seen in \autoref{fig:eq_prf}.
As expected, we have inference rules that utilise the reflexivity, symmetry, and transitivity of equality as well as an assumption rule.
To improve readability, we use the infix operator $\bigtriangledown$ to denote the proof term for transitivity.

@@

definition explain_partial us x y =
  if (x, y) $\in$ equivcl (set us) then Some (explain us x y) else None
  \end{lstlisting}
  \caption{A simple implementation of the \lstinline|explain| operation. It operates on an arbitrary initial \acrshort{uf} data structure \lstinline|uf_init|.\label{fig:explain}}
\end{figure*}
\todo[inline]{termination of \lstinline|explain|}
Although \lstinline|explain| takes a list of equations \lstinline|us|, its implementation actually works with an instance \lstinline|uf| of the \acrshort{uf} \acrshort{adt} that results from applying \lstinline|us| as unions to an initial \acrshort{uf} data structure \lstinline|uf_init|.
Accordingly, we define
\begin{lstlisting}
definition ufa_unions =
  foldl ($\lambda$uf (x, y). ufa_union uf x y)
\end{lstlisting}
so we have \lstinline|uf = ufa_unions uf_init us| in terms of the constants above.

At last, we are ready to implement the \lstinline|explain| operation as depicted in \autoref{fig:explain}.
The algorithm works with the assumption that the given elements \lstinline|x| and \lstinline|y| are equal modulo \lstinline|us|.
Furthermore, we assume that each element of the initial data structure \lstinline|uf_init| is equal to itself
and that \lstinline|us| are valid unions with respect to \lstinline|uf_init|.
In formulae, we assume
\begin{itemize}
  \item \lstinline|ufa_$\alpha$ uf_init $\subseteq$ Id|, and
  \item \lstinline|valid_unions uf_init us|
\end{itemize}
in the remainder of this section.

@@

Otherwise, the list of unions must be equal to \lstinline|us @ [(a, b)]| for some \lstinline|us|, \lstinline|a|, and \lstinline|b|.
We distinguish between two cases:
\begin{itemize}
  \item The elements \lstinline|x| and \lstinline|y| are already equal modulo \lstinline|us|, so we proceed recursively with \lstinline|us|.
  \item In the case that the equation \lstinline|a = b| is necessary for \lstinline|x = y| to hold, we either have \lstinline|x = a| and \lstinline|b = y| or \lstinline|x = b| and \lstinline|a = y| modulo \lstinline|us|.
    Assuming that the former holds --- the other case is symmetric --- we recursively construct the certificates for \lstinline|x = a| and \lstinline|b = y|.
    Together with the assumption that \lstinline|a = b|, we obtain \lstinline|x = y| by transitivity.
\end{itemize}

@@

\section{Efficient Certifying Union-Find Algorithm}
In the previous section, we developed a simple certifying \acrfull{ufe} data structure whose \lstinline|explain| operation iterates through a list of equalities,
identifying which of them are necessary in order to prove that the inputs are equal.
Nevertheless, iterating through all equalities seems inefficient and indeed \citeauthor{congcl_proofs}~\cite{congcl_proofs} propose a more focused approach.
Specifically, they use a \acrshort{uf} data structure represented as forest of rooted trees as described in \autoref{sec:uf_background}.
They modify the data structure such that, for each union between \lstinline|a| and \lstinline|b|, the newly added edge gets annotated with \lstinline|(a, b)|.\todo{Direction of union! Explain sides of the union!}
To understand why this allows for a more efficient implementation of the explain operation, suppose that we want to certify that \lstinline|x| is equal to \lstinline|y|.
Clearly, only the edges of the subtree rooted at the \acrfull{lca} of \lstinline|x| and \lstinline|y|, as illustrated in \autoref{fig:explain'_abs}, are relevant to explain why \lstinline|x| is equal to \lstinline|y|.
Furthermore, let \lstinline|(a, b)| be the most recent union on either of the paths between the \acrshort{lca} and \lstinline|x| and \lstinline|y|, respectively.
\todo[inline]{Intuition: path from x to y}
Here, we assume w.l.o.g.\ that \lstinline|(a, b)| is on the path to \lstinline|x|.
The corresponding edge separates the tree rooted at the \acrshort{lca} into two subtrees as indicated by the patterns,  one containing \lstinline|a| and the other one \lstinline|b|.
Moreover, the paths from the \acrshort{lca} can't overlap, so \lstinline|x| and \lstinline|y| also belong to different subtrees.
Ultimately, to certify the equality of \lstinline|x| and \lstinline|y|, we recursively prove that \lstinline|x| is equal to \lstinline|a| and \lstinline|y| to \lstinline|b|.
Then, we put everything together using transitivity and the equality of \lstinline|a| and \lstinline|b|.
This terminates since \lstinline|(a, b)| is the most recent union and we only consider less recent unions in the recursive steps.
All in all, this gives a $\mathcal{O}(k \log n)$ explain operation where $k$ is the number of unions required for an explanation~\cite{congcl_proofs}.
\todo[inline]{Running time of naive algorithm}
\begin{figure}
  \centering
  \begin{tikzpicture}[
    >=latex, node distance=0.5cm,
    aside/.append style={pattern=north east lines, pattern color=gray!70},
    bside/.append style={pattern={Dots[radius=0.85pt]}, pattern color=gray!70}
    ]
    \node[draw, circle, preaction={fill, white}, style=bside] (lca) {\acrshort{lca}};
    \begin{pgfonlayer}{background}
      \begin{scope}[shape=isosceles triangle, shape border rotate=90, minimum height=2cm, minimum width=2.25cm]
        \node[draw, anchor=north, below left=2.5cm and 3.5cm of lca.south west, aside] (l) {};
        \path (l.north) -- node[pos=0.54, draw, anchor=north, bside] (m) {} (lca.west);
        \node[draw, anchor=north, yshift=0.25cm, bside] (r) at (lca.south) {};
      \end{scope}
    \end{pgfonlayer}

    \node[draw, circle, above right=0.2cm and 0.05cm of l.south west] (x) {\lstinline|x|};
    \node[draw, circle, left=0.1cm of l.east] (a) {\lstinline|a|};
    \node[draw, circle, above left=of r.south east] (y) {\lstinline|y|};
    \node[draw, circle, above right=of m.south west] (b) {\lstinline|b|};

    \draw[->] (l.north) -- node[above, sloped] {\lstinline|(a,$\,$b)|} (m.north);
    \draw[->, dashed] (m.north) -- (lca.west);
    \draw[solid] (lca.north) -- ++(0,0.3);
    \draw[dashed] (lca.north) ++(0,0.3) -- ++(0,0.7);
  \end{tikzpicture}
  \caption{High-level workings of the \lstinline|explain'| function.\label{fig:explain'_abs}}
\end{figure}
